---
output: html_document
---

```{r echo = FALSE}
input_file <- "C:/Users/JV/Desktop/R - Springboard/DallasCrimeData/Police_Incidents.tsv"
output_file <- "C:/Users/JV/Desktop/R - Springboard/DallasCrimeData/Police_Incidents_clean.csv"
```

# **Data Wrangling on Dallas Crime data**

```{r include = FALSE}
library(dplyr)
library(tidyr)
library(lubridate)
```

We use the read.delim() function to read the crime data and store it in a dataframe. We can either read the data directly from the URL or download and read it from the file. Here, we are using the second approach to read the data in R.

We use the read.delim() function to read the crime data and store it in a dataframe. We can either read the data directly from the URL or download and read it from the file. Here, we are using the second approach to read the data in R.

```{r echo = TRUE, warning = FALSE, message = FALSE}
library(dplyr)
library(tidyr)
library(lubridate)
library(kableExtra)
```

```{r echo = TRUE}
crimedata <- read.delim(input_file, header = TRUE,sep = "\t", check.names = FALSE, stringsAsFactors = FALSE)
```

Now that we have read the data, let us understand the data, its fields and its types. str() function displays the structure of the data in a compact way.

```{r echo = TRUE}
str(crimedata)
```

The summary() function provides the detailed summary of data.
```{r echo = FALSE, eval = FALSE}
summary(crimedata)
```

Looking at the summary of data, we understand that there is one observation for each crime incident in the data frame. We have ```{r echo = FALSE} nrow(crimedata) ``` (rows) of 100 variables (columns) where each row is a crime incident reported to the Dallas Police Department. For the ease of data analysis, we select only the fields that are neccessary. 

```{r echo = TRUE}
crimedata <- crimedata %>% select(`Incident Number w/year`,`Year of Incident`,`Type of Incident`,Beat,Division, `Type of Property`,`Date1 of Occurrence`,`Month1 of Occurence`,`Time1 of Occurrence`,`Day1 of the Week`,`Day1 of the Year`,`Victim Age`,`Victim Gender`, `Offense Status`, `UCR Offense Name`, `Victim Condition`, `Weapon Used`)
```

Now with the selected fields, we can start cleaning the data and make it ready for analysis.

* Incident Number should be unique

Each incident has a unique identifier associated with it which is stored in the variable `Incident number w/ year`. However we have some instances where two or more rows have the same identifier. These duplicated instances should be removed. We use the duplicated() function to remove the duplicates.

```{r echo = TRUE}
crimedata <- crimedata[!duplicated(crimedata$`Incident Number w/year`), ]
```

* Date of Occurrence should be between June 2014 and `r Sys.Date() `

Date of Occurrence is a field that stores the date when the incident was reported. Hence it should be of type *Date*. The values should be from June 2014 to `r Sys.Date() `.

```{r echo = TRUE}
crimedata$`Date1 of Occurrence` <- as.Date(crimedata$`Date1 of Occurrence`, format = "%m/%d/%Y")
crimedata <- crimedata %>% filter(between(`Date1 of Occurrence`, as.Date("2014-06-01"), Sys.Date()))
```

* Year of Incident, Month of Occurrence, Day of the Week, Day of the Year should be derived from `Date of Occurrence`

It makes much sense to have the Year, Month, Day of week and Day of year values to be derived from the fields `Date of Occurrence`. This can be easily done with the *lubridate* package.

```{r echo = TRUE}
crimedata$`Year of Incident` <- with(crimedata, ifelse(`Year of Incident` == year(`Date1 of Occurrence`), `Year of Incident` , year(`Date1 of Occurrence`) ))
crimedata$`Month1 of Occurence` <- with(crimedata, ifelse(`Month1 of Occurence` == month(`Date1 of Occurrence`), `Month1 of Occurence`, month(`Date1 of Occurrence`)))
crimedata$`Day1 of the Year` <- with(crimedata, ifelse(`Day1 of the Year` == yday(`Date1 of Occurrence`), `Day1 of the Year`, yday(`Date1 of Occurrence`)))
crimedata$`Day1 of the Week` <- with(crimedata, ifelse(`Day1 of the Week` == wday(`Date1 of Occurrence`), `Day1 of the Week`, wday(`Date1 of Occurrence`)))
```

* Victim Age - Reset values greater than 125 to 125
```{r echo = TRUE}
crimedata$`Victim Age`[crimedata$`Victim Age` > 125 ] = 125
```

* Victim Age cannot be a negative value. 
```{r echo = TRUE}
crimedata$`Victim Age` <- ifelse(crimedata$`Victim Age`< 0 , abs(crimedata$`Victim Age`) , crimedata$`Victim Age`)
```

The variable `Division` tells us in which part of the Dallas city the crime incident took place. It is categorized into 7 divisions - CENTRAL, NORTHEAST, SOUTH CENTRAL, SOUTHWEST, NORTH CENTRAL, NORTHWEST, SOUTHEAST. Therefore, it has to be of type *Factor*

```{r echo = TRUE}
unique(crimedata$Division)
```

It's clear that the values here are not unique, hence converting them all to Upper case help make the analysis easy. 

* Division - Change all the values to Upper case 
```{r echo = TRUE}
tmp <- crimedata %>% mutate(new_division = toupper(Division))
crimedata$Division <- tmp$new_division
crimedata$Division <- as.factor(crimedata$Division)
```

In order to find out how the crime rate trends throughout the year, we create a new varaiable named `Season` based on the `Month of Occurence`. It takes the values - *Spring*, *Summer*, *Fall*, *Winter* and is of type *Factor*.

* Find the Season of the Year i.e., Spring/Summer/Fall/Winter
```{r echo = TRUE}
crimedata <- crimedata %>% mutate(Season = ifelse(`Month1 of Occurence` %in% c(3,4,5), "Spring", 
                                     ifelse(`Month1 of Occurence`%in% c(6,7,8), "Summer",
                                           ifelse(`Month1 of Occurence` %in% c(9,10,11), "Fall",
                                                  ifelse(`Month1 of Occurence` %in% c(12,1,2), "Winter", NA)))))
crimedata$Season <- as.factor(crimedata$Season)
```

* Columns - Victim Gender, Victim Condition, Offense Status should of type *Factor*
```{r echo = TRUE}
crimedata$`Victim Gender` <- as.factor(crimedata$`Victim Gender`)
crimedata$`Victim Condition` <- as.factor(crimedata$`Victim Condition`)
crimedata$`Offense Status` <- as.factor(crimedata$`Offense Status`)
```

We can also try to find out the time of day most crimes tend to happen. For this, we extract the hour of the day from the `Time of Occurrence` variable.

* Extract the hour of the day when the crime took place.
```{r echo = TRUE}
crimedata <- crimedata %>% mutate(`Hour of the Day` = sub(":.*", "",`Time1 of Occurrence`))
```

Variable `Type of Property` stores the target item of the incident. Example : Motor vehicle, Apartment. It cannot take numeric values. 

* `Type of Property` cannot have numeric values
```{r echo = TRUE}
crimedata$`Type of Property` <- as.factor(crimedata$`Type of Property`)
crimedata$`Type of Property` <- droplevels(crimedata$`Type of Property`, exclude = c(910,920,932,510))
```
`UCR Offense Name` stores the type of crime incident that took place. For all the crimes that happened in the year 2019, there is no value for `UCR Offense Name`. This can be extracted from the column `Type of Incident`.

To do this, we create a dataframe which maps the unique `Type of Incident` to its corresponding `UCR Offense Name` called *offenseNames*. Based on this mapping, we find all the missing values for the column `UCR Offense Names`.

```{r echo = TRUE}
temp <- crimedata[!duplicated(crimedata$`Type of Incident`), ]
offenseNames <- temp %>% dplyr::select(`Type of Incident`, `UCR Offense Name`)

offenseNames$`UCR Offense Name`[which(startsWith(offenseNames$`Type of Incident`, "ARSON"))] <- "ARSON"
offenseNames$`UCR Offense Name`[which(grepl(paste(c("^GRAFFITI", "^CRUELTY TO", "^CRIM MISCHIEF"), collapse="|"),offenseNames$`Type of Incident` ))] <- "VANDALISM & CRIM MISCHIEF"
offenseNames$`UCR Offense Name`[which(grepl(paste(c("^ASSAULT", "^DEADLY CONDUCT"), collapse="|"),offenseNames$`Type of Incident` ))] <- "ASSAULT"
offenseNames$`UCR Offense Name`[which(startsWith(offenseNames$`Type of Incident`, "BMV"))] <- "THEFT/BMV"
offenseNames$`UCR Offense Name`[which(grepl(paste(c("^CREDIT CARD", "^COMPUTER SECURITY", "^DECEPTIVE", "^FRAUD", "^THEFT OF SERVICE", "^FALSE STATEMENT", "^TAMPER W" , "^SECURE EXE", "^FAIL TO", "FALSE ALARM"), collapse="|"),offenseNames$`Type of Incident` ))] <- "FRAUD"
offenseNames$`UCR Offense Name`[which(startsWith(offenseNames$`Type of Incident`, "CRIMINAL TRESPASS"))] <- "CRIMINAL TRESPASS"
offenseNames$`UCR Offense Name`[which(grepl(paste(c("^DELIVERY", "^MAN DEL", "^POSS CONT", "^POSS MARIJUANA"), collapse="|"),offenseNames$`Type of Incident` ))] <- "NARCOTICS & DRUG"
offenseNames$`UCR Offense Name`[which(grepl(paste(c("^DISORDERLY", "^DISRUPT", "^ILLUMINA", "^ONLINE IMPRESS", "^STALKING", "^SEX OFFENDERS", "^HARASSMENT"), collapse="|"),offenseNames$`Type of Incident` ))] <- "DISORDERLY CONDUCT"
offenseNames$`UCR Offense Name`[which(startsWith(offenseNames$`Type of Incident`, "DWI"))] <- "DWI"
offenseNames$`UCR Offense Name`[which(startsWith(offenseNames$`Type of Incident`, "ESCAPE"))] <- "ESCAPE"
offenseNames$`UCR Offense Name`[which(startsWith(offenseNames$`Type of Incident`, "EVADING"))] <- "EVADING"
offenseNames$`UCR Offense Name`[which(startsWith(offenseNames$`Type of Incident`, "FORGERY"))] <- "FORGE & COUNTERFEIT"
offenseNames$`UCR Offense Name`[which(startsWith(offenseNames$`Type of Incident`, "KIDNAPPING"))] <- "KIDNAPPING"
offenseNames$`UCR Offense Name`[which(grepl(paste(c("^ILLEGAL", "^INTERFERE", "^INTERFER", "^FLEEING", "^MISAPP", "^OTHER OFFENSES", "^WARRANT"), collapse="|"),offenseNames$`Type of Incident` ))] <- "OTHERS"
offenseNames$`UCR Offense Name`[which(startsWith(offenseNames$`Type of Incident`, "TRAFFICKING"))] <- "HUMAN TRAFFICKING"
offenseNames$`UCR Offense Name`[which(startsWith(offenseNames$`Type of Incident`, "UNAUTHORIZED USE OF"))] <- "UUMV"
offenseNames$`UCR Offense Name`[which(grepl(paste(c("^UNLAWFULLY", "^UNLAWFUL", "^PROHIBITED"), collapse="|"),offenseNames$`Type of Incident` ))] <- "WEAPONS"
offenseNames$`UCR Offense Name`[which(grepl(paste(c("^VIO BOND", "^VIO PROTECT"), collapse="|"),offenseNames$`Type of Incident` ))] <- "OFFENSE AGAINST CHILD"
offenseNames$`UCR Offense Name`[which(grepl(paste(c("^TRAFFIC VIO", "^TRAF VIO"), collapse="|"),offenseNames$`Type of Incident` ))] <- "TRAFFIC VIOLATION"
offenseNames$`UCR Offense Name`[which(startsWith(offenseNames$`Type of Incident`, "TRADEMARK"))] <- "FORGE & COUNTERFEIT"
offenseNames$`UCR Offense Name`[which(startsWith(offenseNames$`Type of Incident`, "THEFT ORG"))] <- "THEFT ORG RETAIL"
offenseNames$`UCR Offense Name`[which(startsWith(offenseNames$`Type of Incident`, "TERRORISTIC THREAT"))] <- "TERRORISTIC THREAT"
offenseNames$`UCR Offense Name`[which(startsWith(offenseNames$`Type of Incident`, "RESIST ARREST"))] <- "RESIST ARREST"
offenseNames$`UCR Offense Name`[which(grepl(paste(c("^MURDER", "^MANSLAUGHTER"), collapse="|"),offenseNames$`Type of Incident` ))] <- "MURDER"
offenseNames$`UCR Offense Name`[which(grepl(paste(c("^PUBLIC INTOX", "^PURCHASE FURN"), collapse="|"),offenseNames$`Type of Incident` ))] <- "DRUNK & DISORDERLY"
offenseNames$`UCR Offense Name`[which(grepl(paste(c("^ROBBERY *OF BUSINESS"), collapse="|"),offenseNames$`Type of Incident` ))] <- "ROBBERY-BUSINESS"
offenseNames$`UCR Offense Name`[which(grepl(paste(c("^ROBBERY *OF INDIVIDUAL"), collapse="|"),offenseNames$`Type of Incident` ))] <- "ROBBERY-INDIVIDUAL"
offenseNames$`UCR Offense Name` <- with(offenseNames, ifelse(`UCR Offense Name` == "" & grepl("^THEFT", `Type of Incident`), "OTHER THEFTs" , `UCR Offense Name` ))
offenseNames$`UCR Offense Name`[which(offenseNames$`UCR Offense Name` == "")] <- "OTHERS"

crimedata$`UCR Offense Name` <- offenseNames[match(crimedata$`Type of Incident`, offenseNames$`Type of Incident`),2]
```

```{r echo = TRUE}
length(unique(crimedata$`UCR Offense Name`))
```

There are 50 different types of crime. We can group similar categories of crime into one and make this number smaller.

* Group similar offense types
```{r echo = TRUE}
crimedata$`Crime Type` <- crimedata$`UCR Offense Name`

crimedata$`Crime Type`[which(crimedata$`Crime Type` %in% c("ASSAULT", "AGG ASSAULT - NFV"))] = "ASSAULT"
crimedata$`Crime Type`[which(crimedata$`Crime Type` %in% c("BURGLARY-BUSINESS", "BURGLARY-RESIDENCE"))] = "BURGLARY"
crimedata$`Crime Type`[which(crimedata$`Crime Type` %in% c("THEFT/BMV", "THEFT/SHOPLIFT", "OTHER THEFTS", "THEFT ORG RETAIL", "EMBEZZLEMENT"))] = "THEFT"
crimedata$`Crime Type`[which(crimedata$`Crime Type` %in% c("ROBBERY-BUSINESS", "ROBBERY-INDIVIDUAL"))] = "ROBBERY"
crimedata$`Crime Type`[which(crimedata$`Crime Type` %in% c("ACCIDENT MV", "MOTOR VEHICLE ACCIDENT"))] = "ACCIDENT"
crimedata$`Crime Type`[which(crimedata$`Crime Type` %in% c("NARCOTICS & DRUGS", "NARCOTICS & DRUG" ,"DRUNK & DISORDERLY", "DWI", "LIQUOR OFFENSE", "INTOXICATION MANSLAUGHTER"))] = "DRUGS"
crimedata$`Crime Type`[which(crimedata$`Crime Type` %in% c("TRAFFIC VIOLATION", "TRAFFIC FATALITY"))] = "TRAFFIC"
crimedata$`Crime Type`[which(crimedata$`Crime Type` %in% c("MURDER", "SUDDEN DEATH&FOUND BODIES", "VANDALISM & CRIM MISCHIEF", "WEAPONS", "ARSON", "TERRORISTIC THREAT", "KIDNAPPING", "HUMAN TRAFFICKING", "OFFENSE AGAINST CHILD", "ORANIZED CRIME"))] = "VIOLENCE"
crimedata$`Crime Type`[which(crimedata$`Crime Type` %in% c("DISORDERLY CONDUCT" ,"CRIMINAL TRESPASS", "EVADING", "RESIST ARREST", "FAIL TO ID", "GAMBLING", "ESCAPE", "FRAUD", "UUMV", "FORGE & COUNTERFEIT"))] = "NONVIOLENCE"
crimedata$`Crime Type`[which(crimedata$`Crime Type` %in% c("NOT CODED", "LOST", "ANIMAL BITE", "OTHERS", "FOUND", "INJURED FIREARM", "INJURED HOME", "INJURED OCCUPA", "INJURED PUBLIC"))] = "OTHERS"
crimedata$`Crime Type` <- as.factor(crimedata$`Crime Type`)
```

Similarly, let us group all the similar weapon categories and store it as *Factor*

```{r echo = TRUE}
crimedata$`Weapon Type` <- ""
crimedata$`Weapon Type`[which(grepl("gun",crimedata$`Weapon Used`, ignore.case = TRUE))] <- "Gun"
crimedata$`Weapon Type`[which(crimedata$`Weapon Used` %in% c("Rifle", "Missile/Rock"))] = "Gun"
crimedata$`Weapon Type`[which(crimedata$`Weapon Used` %in% c("Hands-Feet"))] = "Hands/Feet"
crimedata$`Weapon Type`[which(crimedata$`Weapon Used` %in% c("Vehicle", "MOTOR VEHICLE"))] = "Vehicle"
crimedata$`Weapon Type`[which(crimedata$`Weapon Used` %in% c("None"))] = "No Weapons"
crimedata$`Weapon Type`[which(crimedata$`Weapon Used` %in% c("Threats"))] = "Threat"
crimedata$`Weapon Type`[which(grepl("knife",crimedata$`Weapon Used`, ignore.case = TRUE))] = "Knife"
crimedata$`Weapon Type`[which(crimedata$`Weapon Used` %in% c("Other Cutting Stabbing Inst.", "SWITCHBLADE", "AXE", "ICE PICK"))] = "Knife"
crimedata$`Weapon Type`[which(grepl("fire",crimedata$`Weapon Used`, ignore.case = TRUE))] = "Fire"
crimedata$`Weapon Type`[which(crimedata$`Weapon Used` %in% c("Explosives", "Gas/Carbon Monoxide", "Burn/Scald"))] = "Knife"
crimedata$`Weapon Type`[which(grepl("drugs",crimedata$`Weapon Used`, ignore.case = TRUE))] = "Drugs"
crimedata$`Weapon Type`[which(crimedata$`Weapon Used` %in% c("ANY WEAPON OF FORCE DEADLY DISEASE, ETC", "Omission/Neglect"))] = "Drugs"
crimedata$`Weapon Type`[which(crimedata$`Weapon Used` %in% c("Other", "Blunt", "Stangulation", "Assault", "Crowbar", "Asphixation", "BlackJack/Club", "Omission"))] <- "Others"
crimedata$`Weapon Type`[which(crimedata$`Weapon Type` == "")] <- "Others"
crimedata$`Weapon Type` <- as.factor(crimedata$`Weapon Type`)
```


### Write the cleaned dataset to a file.
write.csv(crimedata, output_file, row.names = FALSE)